---
title: "Bioinformatics data skills workshop - Session 1: getting started on the Linux command line"
author: "Burnet Bioinformatics Group"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: simplex
    toc: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
    toc_float:
      collapsed: true
---

Source: TBA

## Introduction

Dual RNA-seq of Aedes aegypti infected with different Wolbachia strains which provide different levels of antiviral action.

Sample groups:

* Tet: this is the non-infected control

* wPip: no antiviral action

* wMel: antiviral

* wAlb: antiviral

```{r,libs}

library("RColorBrewer")
library("DESeq2")
library("beeswarm")
library("gplots")
library("kableExtra")
library("mitch")
library("fgsea")

```

## Load data

Load in sample sheet and expression counts.

```{r,loaddata}

ss <- read.table("samplesheet.tsv",header=TRUE)

ss$SampleGroup <- sapply(strsplit(ss$Original_ID,"_"),"[[",1)

ss$Batch <- sapply(strsplit(unlist(sapply(strsplit(ss$Original_ID,"_"),"[[",2)),"-"),"[[",1)

ss

x <- read.csv("merged_gene_count_matrix.csv",header=TRUE, row.names=1)

head(x)

```

## Fix the colnames

This allows us to correct batches better.

```{r,colnames1}

colnames(x) <- ss$Original_ID[match(colnames(x), ss$Generic_ID_in_R)]

head(x)

```

## Quality control

Check depth.

```{r,qc1}


colSums(x)

par(mar=c(5.1, 8.1, 4.1, 2.1))

barplot(colSums(x),horiz=TRUE,las=1)

par(mar=c(5.1, 4.1, 4.1, 2.1))

```

## Colours

```{r,cols1}

brewer.pal(n = 12, name = "Paired")

display.brewer.pal(n = 12, name = "Paired")

```

## PCA

```{r,mds1}

mds <- cmdscale(dist(t(x)))

par(mar=c(5,5,3,1))
minx <- min(mds[,1])
maxx <- max(mds[,1])
miny <- min(mds[,2])
maxy <- max(mds[,2])

#tet #A6CEE3
#pip #B2DF8A
#wmel #FB9A99
#walb #FDBF6F
cols <- gsub("wAlb", "#FDBF6F ", gsub("wMel", "#FB9A99 ", gsub("wPip", "#B2DF8A ",  gsub("Tet","#A6CEE3 ",colnames(x)) ) ) )

cols <- sapply(strsplit(cols," "),"[[",1)

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(minx*1.1,maxx*1.1), ylim = c(miny*1.1,maxy*1.1) , main="MDS plot",
  type = "p", col=cols, pch=19, cex.axis=1.3, cex.lab=1.3, bty='n', cex=2)
text(mds, labels=rownames(mds), cex=1.2)

```

## Differential analysis

Here we'll make the following comparisons:

1. Tet (ctrl) vs wPip (trt): effect of infection (no antiviral activity)

2. Tet (ctrl) vs wMel (trt): effect of infection with an antiviral strain

3. Tet (ctrl) vs wAlb (trt): effect of infection with an antiviral strain

4. wPip (ctrl) vs wMel (trt): direct comparison of non-antiviral and antiviral

5. wPip (ctrl) vs wAlb (trt): direct comparison of non-antiviral and antiviral

```{r,plots}

volcanoplot <- function(x,title) {
  tot <- nrow(x)
  sig <- subset(x,padj<0.01)
  nsig <- nrow(subset(x,padj<0.01))
  dn <- nrow(subset(x,padj<0.01 & log2FoldChange<0))
  up <- nrow(subset(x,padj<0.01 & log2FoldChange>0))
  header <- paste(title,":", tot, "genes,", nsig ,"@1% FDR,", up, "up,", dn, "dn")

  plot(x$log2FoldChange,-log10(x$pvalue),bty="n",
    cex=0.5, pch=19, xlab="log2 fold change",
    ylab="-log10 p-value", main=header)

  points(sig$log2FoldChange,-log10(sig$pvalue),
    cex=0.5, pch=19, col="red")
}

smearplot <- function(x,title) {
  tot <- nrow(x)
  sig <- subset(x,padj<0.01)
  nsig <- nrow(subset(x,padj<0.01))
  dn <- nrow(subset(x,padj<0.01 & log2FoldChange<0))
  up <- nrow(subset(x,padj<0.01 & log2FoldChange>0))
  header <- paste(title,":", tot, "genes,", nsig ,"@1% FDR,", up, "up,", dn, "dn")

  plot(log10(x$baseMean),x$log2FoldChange,bty="n",
    cex=0.5, pch=19, xlab="log10 base mean",
    ylab="log2 fold change", main=header)

  points(log10(sig$baseMean),sig$log2FoldChange,
    cex=0.5, pch=19, col="red")
}

# summarise a bunch of DE object results
debar <- function(l) {
  nn <- sapply(l,function(x) {
    nup <- nrow(subset(x,padj<0.05 & log2FoldChange>0 ))
    ndn <- nrow(subset(x,padj<0.05 & log2FoldChange<0 )) *-1
    return(c(nup,ndn))
  })
  rownames(nn) <- c("up","dn")
  par(mar = c(8.1, 4.1, 4.1, 2.1))
  barplot(nn[1,],col="blue",beside = TRUE, border=TRUE,ylim=c(min(nn),max(nn)) , las=2)
  barplot(nn[2,],add=TRUE,col="red",names.arg = "", ylab="", yaxt="n")
  mtext("blue=up,red=down")
  par(mar = c(5.1, 4.1, 4.1, 2.1))
}

```



### DE1

Effect of infection (no antiviral activity).

```{r,de1}

ss1 <- subset(ss,SampleGroup=="Tet"|SampleGroup=="wPip")
batch_include <- names(which(table(unique(ss1[,3:4])$Batch)==2))
ss1 <- ss1[which(ss1$Batch %in% batch_include),]
ss1$trt <- grepl("wPip",ss1$SampleGroup)
x1 <- x[,which(colnames(x) %in% ss1$Original_ID)]
table(rowMeans(x1)>10)
x1 <- x1[which(rowMeans(x1)>10),]
dim(x1)

mds <- cmdscale(dist(t(x1)))
par(mar=c(5,5,3,1))
minx <- min(mds[,1])
maxx <- max(mds[,1])
miny <- min(mds[,2])
maxy <- max(mds[,2])
cols <- gsub("wAlb", "#FDBF6F ", gsub("wMel", "#FB9A99 ", gsub("wPip", "#B2DF8A ",  gsub("Tet","#A6CEE3 ",colnames(x1)) ) ) )
cols <- sapply(strsplit(cols," "),"[[",1)

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(minx*1.1,maxx*1.1), ylim = c(miny*1.1,maxy*1.1) , main="MDS plot",
  type = "p", col=cols, pch=19, cex.axis=1.3, cex.lab=1.3, bty='n', cex=2)
text(mds, labels=rownames(mds), cex=1.2)

dds <- DESeqDataSetFromMatrix(countData = x1 , colData = ss1, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
table(is.na(dge$pvalue))
dge <- dge[which(!is.na(dge$pvalue)),]
dim(dge)

head(dge[order(dge$pvalue),1:6],10) %>%
  kbl(caption="Top DE genes for Pip infection") %>%
  kable_paper("hover", full_width = F)

volcanoplot(dge,"DGE1")

smearplot(dge,"DGE1")

dge1 <- dge

```

### DE2

Tet (ctrl) vs wMel (trt): effect of infection with an antiviral strain

```{r,de2}

ss2 <- subset(ss,SampleGroup=="Tet"|SampleGroup=="wMel")
batch_include <- names(which(table(unique(ss2[,3:4])$Batch)==2))
ss2 <- ss2[which(ss2$Batch %in% batch_include),]
ss2$trt <- grepl("wMel",ss2$SampleGroup)
x2 <- x[,which(colnames(x) %in% ss2$Original_ID)]
table(rowMeans(x2)>10)
x2 <- x2[which(rowMeans(x2)>10),]
dim(x2)

mds <- cmdscale(dist(t(x2)))
par(mar=c(5,5,3,1))
minx <- min(mds[,1])
maxx <- max(mds[,1])
miny <- min(mds[,2])
maxy <- max(mds[,2])
cols <- gsub("wAlb", "#FDBF6F ", gsub("wMel", "#FB9A99 ", gsub("wPip", "#B2DF8A ",  gsub("Tet","#A6CEE3 ",colnames(x1)) ) ) )
cols <- sapply(strsplit(cols," "),"[[",1)

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(minx*1.1,maxx*1.1), ylim = c(miny*1.1,maxy*1.1) , main="MDS plot",
  type = "p", col=cols, pch=19, cex.axis=1.3, cex.lab=1.3, bty='n', cex=2)
text(mds, labels=rownames(mds), cex=1.2)

dds <- DESeqDataSetFromMatrix(countData = x2 , colData = ss2, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
table(is.na(dge$pvalue))
dge <- dge[which(!is.na(dge$pvalue)),]
dim(dge)

head(dge[order(dge$pvalue),1:6],10) %>%
  kbl(caption="Top DE genes for wMel infection") %>%
  kable_paper("hover", full_width = F)

volcanoplot(dge,"DGE2")

smearplot(dge,"DGE2")

dge2 <- dge

```

### DE3

Tet (ctrl) vs wAlb (trt): effect of infection with an antiviral strain

```{r,de3}

ss3 <- subset(ss,SampleGroup=="Tet"|SampleGroup=="wAlb")
batch_include <- names(which(table(unique(ss3[,3:4])$Batch)==2))
ss3 <- ss3[which(ss3$Batch %in% batch_include),]
ss3$trt <- grepl("wAlb",ss3$SampleGroup)
x3 <- x[,which(colnames(x) %in% ss3$Original_ID)]
table(rowMeans(x3)>10)
x3 <- x3[which(rowMeans(x3)>10),]
dim(x3)

mds <- cmdscale(dist(t(x3)))
par(mar=c(5,5,3,1))
minx <- min(mds[,1])
maxx <- max(mds[,1])
miny <- min(mds[,2])
maxy <- max(mds[,2])
cols <- gsub("wAlb", "#FDBF6F ", gsub("wMel", "#FB9A99 ", gsub("wPip", "#B2DF8A ",  gsub("Tet","#A6CEE3 ",colnames(x1)) ) ) )
cols <- sapply(strsplit(cols," "),"[[",1)

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(minx*1.1,maxx*1.1), ylim = c(miny*1.1,maxy*1.1) , main="MDS plot",
  type = "p", col=cols, pch=19, cex.axis=1.3, cex.lab=1.3, bty='n', cex=2)
text(mds, labels=rownames(mds), cex=1.2)

dds <- DESeqDataSetFromMatrix(countData = x3 , colData = ss3, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
table(is.na(dge$pvalue))
dge <- dge[which(!is.na(dge$pvalue)),]
dim(dge)

head(dge[order(dge$pvalue),1:6],10) %>%
  kbl(caption="Top DE genes for wMel infection") %>%
  kable_paper("hover", full_width = F)

volcanoplot(dge,"DGE3")

smearplot(dge,"DGE3")

dge3 <- dge

```
