---
title: "Dual RNA-seq impact of wolbachia infection on Aedes"
author: "Fraser Lab and Burnet Bioinformatics Group"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: simplex
    toc: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
    toc_float:
      collapsed: true
---

Source: TBA

## Introduction

Dual RNA-seq of Aedes aegypti infected with different Wolbachia strains which provide different levels of antiviral action.

Sample groups:

* Tet: this is the non-infected control

* wPip: no antiviral action

* wMel: antiviral

* wAlb: antiviral

```{r,libs}

library("RColorBrewer")
library("DESeq2")
library("beeswarm")
library("gplots")
library("kableExtra")
library("mitch")
library("fgsea")
library("eulerr")

```

## Load data

Load in sample sheet and expression counts.

```{r,loaddata}

ss <- read.table("samplesheet.tsv",header=TRUE)

ss$SampleGroup <- sapply(strsplit(ss$Original_ID,"_"),"[[",1)

ss$Batch <- sapply(strsplit(unlist(sapply(strsplit(ss$Original_ID,"_"),"[[",2)),"-"),"[[",1)

ss

x <- read.csv("merged_gene_count_matrix.csv",header=TRUE, row.names=1)

head(x)

```

## Fix the colnames

This allows us to correct batches better.

```{r,colnames1}

colnames(x) <- ss$Original_ID[match(colnames(x), ss$Generic_ID_in_R)]

head(x)

```

## Quality control

Check depth.

```{r,qc1}


colSums(x)

par(mar=c(5.1, 8.1, 4.1, 2.1))

barplot(colSums(x),horiz=TRUE,las=1)

par(mar=c(5.1, 4.1, 4.1, 2.1))

```

## Get gene annotations

The following code was run to get the GO annotations.

```
python getannos.py --ids locs.txt --out go_annotations.csv
```

It mines Aedes data from this NCBI file: "https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene2go.gz"
which was obtained 1st Sept 2026.
Coverage is OK, with 8950 genes having one or more GO annotations.
But there are only 40,915 total annotations, so this is not comprehensive.
Each gene has a median of 3 annotations.
Each GO category has a median of 2 genes.

```{r,gene_anno1}

go <- read.csv("go_annotations.csv")
head(go)
summary(as.numeric(table(go$GO_ID)))
hist(as.numeric(table(go$GO_ID)))

go_sets <- lapply(unique(go$GO_ID),function(x) { go[which(go$GO_ID==x),"InputID"] } )

names(go_sets) <- unlist(lapply(unique(go$GO_ID),function(x) {
  goterm <- go[which(go$GO_ID==x)[1],"GO_term"]
  qualifier <-  go[which(go$GO_ID==x)[1],"Qualifier"]
  y <- paste(x,qualifier,goterm)
  return(y)
} ) )

length(rownames(x))
table(rownames(x) %in% go$InputID)

```

## Colours

```{r,cols1}

brewer.pal(n = 12, name = "Paired")

display.brewer.pal(n = 12, name = "Paired")

```

## PCA

```{r,mds1}

mds <- cmdscale(dist(t(x)))

par(mar=c(5,5,3,1))
minx <- min(mds[,1])
maxx <- max(mds[,1])
miny <- min(mds[,2])
maxy <- max(mds[,2])

#tet #A6CEE3
#pip #B2DF8A
#wmel #FB9A99
#walb #FDBF6F
cols <- gsub("wAlb", "#FDBF6F ", gsub("wMel", "#FB9A99 ", gsub("wPip", "#B2DF8A ",  gsub("Tet","#A6CEE3 ",colnames(x)) ) ) )

cols <- sapply(strsplit(cols," "),"[[",1)

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(minx*1.1,maxx*1.1), ylim = c(miny*1.1,maxy*1.1) , main="MDS plot",
  type = "p", col=cols, pch=19, cex.axis=1.3, cex.lab=1.3, bty='n', cex=2)
text(mds, labels=rownames(mds), cex=1.2)

```

## Correlation analysis

```{r,cor1}

mycor <- cor(x)
heatmap.2( mycor,scale="none", trace="none",margins = c(6,6), cexRow=1, main="Pearson Correlation")

```

## Differential analysis

Here we'll make the following comparisons:

1. Tet (ctrl) vs wPip (trt): effect of infection (no antiviral activity)

2. Tet (ctrl) vs wMel (trt): effect of infection with an antiviral strain

3. Tet (ctrl) vs wAlb (trt): effect of infection with an antiviral strain

4. wPip (ctrl) vs wMel (trt): direct comparison of non-antiviral and antiviral

5. wPip (ctrl) vs wAlb (trt): direct comparison of non-antiviral and antiviral

```{r,plots}

volcanoplot <- function(x,title) {
  tot <- nrow(x)
  sig <- subset(x,padj<0.01)
  nsig <- nrow(subset(x,padj<0.01))
  dn <- nrow(subset(x,padj<0.01 & log2FoldChange<0))
  up <- nrow(subset(x,padj<0.01 & log2FoldChange>0))
  header <- paste(title,":", tot, "genes,", nsig ,"@1% FDR,", up, "up,", dn, "dn")

  plot(x$log2FoldChange,-log10(x$pvalue),bty="n",
    cex=0.5, pch=19, xlab="log2 fold change",
    ylab="-log10 p-value", main=header)

  points(sig$log2FoldChange,-log10(sig$pvalue),
    cex=0.5, pch=19, col="red")
}

smearplot <- function(x,title) {
  tot <- nrow(x)
  sig <- subset(x,padj<0.01)
  nsig <- nrow(subset(x,padj<0.01))
  dn <- nrow(subset(x,padj<0.01 & log2FoldChange<0))
  up <- nrow(subset(x,padj<0.01 & log2FoldChange>0))
  header <- paste(title,":", tot, "genes,", nsig ,"@1% FDR,", up, "up,", dn, "dn")

  plot(log10(x$baseMean),x$log2FoldChange,bty="n",
    cex=0.5, pch=19, xlab="log10 base mean",
    ylab="log2 fold change", main=header)

  points(log10(sig$baseMean),sig$log2FoldChange,
    cex=0.5, pch=19, col="red")
}

# summarise a bunch of DE object results
debar <- function(l) {
  nn <- sapply(l,function(x) {
    nup <- nrow(subset(x,padj<0.05 & log2FoldChange>0 ))
    ndn <- nrow(subset(x,padj<0.05 & log2FoldChange<0 )) *-1
    return(c(nup,ndn))
  })
  rownames(nn) <- c("up","dn")
  par(mar = c(8.1, 4.1, 4.1, 2.1))
  barplot(nn[1,],col="blue",beside = TRUE, border=TRUE,ylim=c(min(nn),max(nn)) , las=2)
  barplot(nn[2,],add=TRUE,col="red",names.arg = "", ylab="", yaxt="n")
  mtext("blue=up,red=down")
  par(mar = c(5.1, 4.1, 4.1, 2.1))
}

```

### DE1

Effect of infection (no antiviral activity).

```{r,de1}

ss1 <- subset(ss,SampleGroup=="Tet"|SampleGroup=="wPip")
batch_include <- names(which(table(unique(ss1[,3:4])$Batch)==2))
ss1 <- ss1[which(ss1$Batch %in% batch_include),]
ss1$trt <- grepl("wPip",ss1$SampleGroup)
x1 <- x[,which(colnames(x) %in% ss1$Original_ID)]
table(rowMeans(x1)>10)
x1 <- x1[which(rowMeans(x1)>10),]
dim(x1)

mds <- cmdscale(dist(t(x1)))
par(mar=c(5,5,3,1))
minx <- min(mds[,1])
maxx <- max(mds[,1])
miny <- min(mds[,2])
maxy <- max(mds[,2])
cols <- gsub("wAlb", "#FDBF6F ", gsub("wMel", "#FB9A99 ", gsub("wPip", "#B2DF8A ",  gsub("Tet","#A6CEE3 ",colnames(x1)) ) ) )
cols <- sapply(strsplit(cols," "),"[[",1)

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(minx*1.1,maxx*1.1), ylim = c(miny*1.1,maxy*1.1) , main="MDS plot",
  type = "p", col=cols, pch=19, cex.axis=1.3, cex.lab=1.3, bty='n', cex=2)
text(mds, labels=rownames(mds), cex=1.2)

dds <- DESeqDataSetFromMatrix(countData = x1 , colData = ss1, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
table(is.na(dge$pvalue))
dge <- dge[which(!is.na(dge$pvalue)),]
dim(dge)

head(dge[order(dge$pvalue),1:6],10) %>%
  kbl(caption="Top DE genes for Pip infection") %>%
  kable_paper("hover", full_width = F)

volcanoplot(dge,"DGE1")

smearplot(dge,"DGE1")

dge1 <- dge

```

### DE2

Tet (ctrl) vs wMel (trt): effect of infection with an antiviral strain

```{r,de2}

ss2 <- subset(ss,SampleGroup=="Tet"|SampleGroup=="wMel")
batch_include <- names(which(table(unique(ss2[,3:4])$Batch)==2))
ss2 <- ss2[which(ss2$Batch %in% batch_include),]
ss2$trt <- grepl("wMel",ss2$SampleGroup)
x2 <- x[,which(colnames(x) %in% ss2$Original_ID)]
table(rowMeans(x2)>10)
x2 <- x2[which(rowMeans(x2)>10),]
dim(x2)

mds <- cmdscale(dist(t(x2)))
par(mar=c(5,5,3,1))
minx <- min(mds[,1])
maxx <- max(mds[,1])
miny <- min(mds[,2])
maxy <- max(mds[,2])
cols <- gsub("wAlb", "#FDBF6F ", gsub("wMel", "#FB9A99 ", gsub("wPip", "#B2DF8A ",  gsub("Tet","#A6CEE3 ",colnames(x2)) ) ) )
cols <- sapply(strsplit(cols," "),"[[",1)

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(minx*1.1,maxx*1.1), ylim = c(miny*1.1,maxy*1.1) , main="MDS plot",
  type = "p", col=cols, pch=19, cex.axis=1.3, cex.lab=1.3, bty='n', cex=2)
text(mds, labels=rownames(mds), cex=1.2)

dds <- DESeqDataSetFromMatrix(countData = x2 , colData = ss2, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
table(is.na(dge$pvalue))
dge <- dge[which(!is.na(dge$pvalue)),]
dim(dge)

head(dge[order(dge$pvalue),1:6],10) %>%
  kbl(caption="Top DE genes for wMel infection") %>%
  kable_paper("hover", full_width = F)

volcanoplot(dge,"DGE2")

smearplot(dge,"DGE2")

dge2 <- dge

```

### DE3

Tet (ctrl) vs wAlb (trt): effect of infection with an antiviral strain

```{r,de3}

ss3 <- subset(ss,SampleGroup=="Tet"|SampleGroup=="wAlb")
batch_include <- names(which(table(unique(ss3[,3:4])$Batch)==2))
ss3 <- ss3[which(ss3$Batch %in% batch_include),]
ss3$trt <- grepl("wAlb",ss3$SampleGroup)
x3 <- x[,which(colnames(x) %in% ss3$Original_ID)]
table(rowMeans(x3)>10)
x3 <- x3[which(rowMeans(x3)>10),]
dim(x3)

mds <- cmdscale(dist(t(x3)))
par(mar=c(5,5,3,1))
minx <- min(mds[,1])
maxx <- max(mds[,1])
miny <- min(mds[,2])
maxy <- max(mds[,2])
cols <- gsub("wAlb", "#FDBF6F ", gsub("wMel", "#FB9A99 ", gsub("wPip", "#B2DF8A ",  gsub("Tet","#A6CEE3 ",colnames(x3)) ) ) )
cols <- sapply(strsplit(cols," "),"[[",1)

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(minx*1.1,maxx*1.1), ylim = c(miny*1.1,maxy*1.1) , main="MDS plot",
  type = "p", col=cols, pch=19, cex.axis=1.3, cex.lab=1.3, bty='n', cex=2)
text(mds, labels=rownames(mds), cex=1.2)

dds <- DESeqDataSetFromMatrix(countData = x3 , colData = ss3, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
table(is.na(dge$pvalue))
dge <- dge[which(!is.na(dge$pvalue)),]
dim(dge)

head(dge[order(dge$pvalue),1:6],10) %>%
  kbl(caption="Top DE genes for wMel infection") %>%
  kable_paper("hover", full_width = F)

volcanoplot(dge,"DGE3")

smearplot(dge,"DGE3")

dge3 <- dge

```


### DE4

wPip (ctrl) vs wMel (trt): direct comparison of non-antiviral and antiviral

```{r,de4}

ss4 <- subset(ss,SampleGroup=="wPip"|SampleGroup=="wMel")
batch_include <- names(which(table(unique(ss4[,3:4])$Batch)==2))
ss4 <- ss4[which(ss4$Batch %in% batch_include),]
ss4$trt <- grepl("wMel",ss4$SampleGroup)
x4 <- x[,which(colnames(x) %in% ss4$Original_ID)]
table(rowMeans(x4)>10)
x4 <- x4[which(rowMeans(x4)>10),]
dim(x4)

mds <- cmdscale(dist(t(x4)))
par(mar=c(5,5,3,1))
minx <- min(mds[,1])
maxx <- max(mds[,1])
miny <- min(mds[,2])
maxy <- max(mds[,2])
cols <- gsub("wAlb", "#FDBF6F ", gsub("wMel", "#FB9A99 ", gsub("wPip", "#B2DF8A ",  gsub("Tet","#A6CEE3 ",colnames(x4)) ) ) )
cols <- sapply(strsplit(cols," "),"[[",1)

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(minx*1.1,maxx*1.1), ylim = c(miny*1.1,maxy*1.1) , main="MDS plot",
  type = "p", col=cols, pch=19, cex.axis=1.3, cex.lab=1.3, bty='n', cex=2)
text(mds, labels=rownames(mds), cex=1.2)

dds <- DESeqDataSetFromMatrix(countData = x4 , colData = ss4, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
table(is.na(dge$pvalue))
dge <- dge[which(!is.na(dge$pvalue)),]
dim(dge)

head(dge[order(dge$pvalue),1:6],10) %>%
  kbl(caption="Top DE genes for wMel infection") %>%
  kable_paper("hover", full_width = F)

volcanoplot(dge,"DGE4")

smearplot(dge,"DGE4")

dge4 <- dge

```

### DE5

wPip (ctrl) vs wAlb (trt): direct comparison of non-antiviral and antiviral

```{r,de5}

ss5 <- subset(ss,SampleGroup=="wPip"|SampleGroup=="wAlb")
batch_include <- names(which(table(unique(ss5[,3:4])$Batch)==2))
ss5 <- ss5[which(ss5$Batch %in% batch_include),]
ss5$trt <- grepl("wAlb",ss5$SampleGroup)
x5 <- x[,which(colnames(x) %in% ss5$Original_ID)]
table(rowMeans(x5)>10)
x5 <- x5[which(rowMeans(x5)>10),]
dim(x5)

mds <- cmdscale(dist(t(x5)))
par(mar=c(5,5,3,1))
minx <- min(mds[,1])
maxx <- max(mds[,1])
miny <- min(mds[,2])
maxy <- max(mds[,2])
cols <- gsub("wAlb", "#FDBF6F ", gsub("wMel", "#FB9A99 ", gsub("wPip", "#B2DF8A ",  gsub("Tet","#A6CEE3 ",colnames(x5)) ) ) )
cols <- sapply(strsplit(cols," "),"[[",1)

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(minx*1.1,maxx*1.1), ylim = c(miny*1.1,maxy*1.1) , main="MDS plot",
  type = "p", col=cols, pch=19, cex.axis=1.3, cex.lab=1.3, bty='n', cex=2)
text(mds, labels=rownames(mds), cex=1.2)

dds <- DESeqDataSetFromMatrix(countData = x5 , colData = ss5, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
table(is.na(dge$pvalue))
dge <- dge[which(!is.na(dge$pvalue)),]
dim(dge)

head(dge[order(dge$pvalue),1:6],10) %>%
  kbl(caption="Top DE genes for wMel infection") %>%
  kable_paper("hover", full_width = F)

volcanoplot(dge,"DGE5")

smearplot(dge,"DGE5")

dge5 <- dge

```


## Correlation analysis

```{r,correl1}

xl <- list("d1"=dge1,"d2"=dge2,"d3"=dge3,"d4"=dge4,"d5"=dge5)
m <- mitch_import(x=xl,DEtype="DEseq2")
mycor2 <- cor(m)
heatmap.2( mycor2, scale="none", trace="none",margins = c(6,6), main="Pearson Correlation")

mycor2 %>%
  kbl(caption="Pearson correlations of contrasts") %>%
  kable_paper("hover", full_width = F)


```

## Set analysis

```{r,set_analysis1}

fdr=0.01
lfc=0.5

eulerplot <- function(x,y,namex,namey) {
  xup <- rownames(subset(x,padj<fdr & log2FoldChange>lfc))
  xdn <- rownames(subset(x,padj<fdr & log2FoldChange<lfc))
  yup <- rownames(subset(y,padj<fdr & log2FoldChange>lfc))
  ydn <- rownames(subset(y,padj<fdr & log2FoldChange<lfc))
  nxup <- paste(namex,"up")
  nxdn <- paste(namex,"dn")
  nyup <- paste(namey,"up")
  nydn <- paste(namey,"dn")
  v1 <- list(nxup=xup, nxdn=xdn, nyup=yup, nydn=ydn)
  names(v1) <- c(nxup, nxdn, nyup, nydn)
  plot(euler(v1),quantities = TRUE)
}

eulerplot(dge1, dge2, "dge1", "dge2")
eulerplot(dge1, dge3, "dge1", "dge3")
eulerplot(dge1, dge4, "dge1", "dge4")
eulerplot(dge1, dge5, "dge1", "dge5")

eulerplot(dge2, dge3, "dge2", "dge3")
eulerplot(dge2, dge4, "dge2", "dge4")
eulerplot(dge2, dge5, "dge2", "dge5")

eulerplot(dge3, dge4, "dge3", "dge4")
eulerplot(dge3, dge5, "dge3", "dge5")

eulerplot(dge4, dge5, "dge4", "dge5")

jaccardindex <- function(x,y) {
  xup <- rownames(subset(x,padj<fdr & log2FoldChange>lfc))
  xdn <- rownames(subset(x,padj<fdr & log2FoldChange<lfc))
  yup <- rownames(subset(y,padj<fdr & log2FoldChange>lfc))
  ydn <- rownames(subset(y,padj<fdr & log2FoldChange<lfc))
  xup <- paste(xup,"up")
  xdn <- paste(xdn,"up")
  yup <- paste(yup,"up")
  ydn <- paste(ydn,"up")
  xde <- c(xup,xdn)
  yde <- c(yup,ydn)
  jac <- length(intersect(xde,yde)) / length(union(xde,yde))
  return(jac)
}

jaccardindex(dge1, dge2)
jaccardindex(dge1, dge3)
jaccardindex(dge1, dge4)
jaccardindex(dge1, dge5)
jaccardindex(dge2, dge3)
jaccardindex(dge2, dge4)
jaccardindex(dge2, dge5)
jaccardindex(dge3, dge4)
jaccardindex(dge3, dge5)
jaccardindex(dge4, dge5)

```

## Select genes specific to antiviral infection

Here we'll make the following comparisons:

1. Tet (ctrl) vs wPip (trt): effect of infection (no antiviral activity)

2. Tet (ctrl) vs wMel (trt): effect of infection with an antiviral strain

3. Tet (ctrl) vs wAlb (trt): effect of infection with an antiviral strain

4. wPip (ctrl) vs wMel (trt): direct comparison of non-antiviral and antiviral

5. wPip (ctrl) vs wAlb (trt): direct comparison of non-antiviral and antiviral

Will get those common to d4 and d5.

```{r,geneselect1}


d1up <- rownames(subset(dge1,padj<fdr & log2FoldChange>lfc))
d1dn <- rownames(subset(dge1,padj<fdr & log2FoldChange<lfc))
d2up <- rownames(subset(dge2,padj<fdr & log2FoldChange>lfc))
d2dn <- rownames(subset(dge2,padj<fdr & log2FoldChange<lfc))

length(intersect(d1up, d2up))
intersect(d1up, d2up)

length(intersect(d1dn, d2dn))
intersect(d1dn, d2dn)

```

## Pathway enrichment 1

DGE1, DGE2, DGE3. One non-antiviral and two antiviral bacterium.

```{r,mitch1}

l1 <- list("cvPip"=dge1,"cvMel"=dge2,"cvAlb"=dge3)
m1 <- mitch_import(x=l1,DEtype="DEseq2")
head(m1)

mres1 <- mitch_calc(x=m1,genesets=go_sets,minsetsize=5,cores=4,priority="effect")
dim(mres1$enrichment_result)
mres1x <- subset(mres1$enrichment_result,p.adjustMANOVA<0.01)

head(mres1x)
pairs(mres1x[,4:6])
mres1x$antiviral <- ( mres1x$s.cvMel + mres1x$s.cvAlb ) / 2
mres1x$SD2 <- apply(mres1x, 1, function(x) { sd(c(x["s.cvPip"] ,x["antiviral"] ) ) } )
mres1x$diff <- mres1x$antiviral - mres1x$s.cvPip
head(mres1x)
mres1x <- mres1x[order(-mres1x$SD2),]

head(mres1x,10) %>%
  kbl(caption="Top GO terms with altered expression in antiviral strains as compared to wPip") %>%
  kable_paper("hover", full_width = F)

message("More common for pathways to be upregulated in antiviral strains")
summary(mres1x$antiviral - mres1x$s.cvPip)
hist(mres1x$antiviral - mres1x$s.cvPip)

mitch_report(res=mres1,outfile="mitch1_report.html",overwrite=TRUE)

```

Now I will focus on these gene sets and make a detailed report.
In this report, we are looking at which GO sets are there and which genes, but the FDR values will
not be correct as we have cherry-picked them based on the above analysis.

```{r,mitch2}

table(names(go_sets) %in% head(mres1x$set,50))

selected_go_sets <- go_sets[which(names(go_sets) %in% head(mres1x$set,50))]

mres2 <- mitch_calc(x=m1,genesets=selected_go_sets,minsetsize=5,cores=4,priority="effect")

mitch_report(res=mres2,outfile="mitch2_report.html",overwrite=TRUE)

```

Make a heatmap of top GO terms.

```{r,goterms,fig.height=8,fig.width=8}

mres2x <- mres2$enrichment_result[,4:6]

rownames(mres2x) <- gsub("_"," ",mres2$enrichment_result$set)

colfunc <- colorRampPalette(c("blue", "white", "red"))

heatmap.2( as.matrix(mres2x), col=colfunc(25), scale="none",
 trace="none",margins = c(5,20), cexRow=.6, cexCol=.8,
 main="GO terms of interest")

```

## Session information

For reproducibility.

```{r,session}

sessionInfo()

```
